---
title: "Week 6: Homework Assignment"
author: "Joshua Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
output:
  html_document:
    theme: united
    df_print: paged
    highlight: textmate
    code_folding: show
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: inline
always_allow_html: yes
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp=.618, fig.align="center",
                      fig.path="Figs/",
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here, rstan, HDInterval, tictoc, furrr, latex2exp, quantmod)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())

# Parallel Stan
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

Consider daily simple returns of AAPL, which, for example, can be obtained as:
```{r}
getSymbols("AAPL", from = "2015-1-1", to = "2015-12-31")
```

The data are in the file `AAPL_2015.csv` 
```{r}
AAPL.1.day <- as.matrix(AAPL)
AAPL.returns.2015 <- diff(AAPL.1.day[, 6])
```

```{r}
AAPL.ret <- read_csv("documents-MScA Bayesian Methods (32014)-MScA 32014 Lecture 7-AAPL_2015.csv")
(AAPL.ret)
```

Visualize the distribution of the returns.
```{r}
data.frame(returns = AAPL.returns.2015) %>%
  ggplot() +
  geom_density(aes(returns)) +
  geom_line(data = data.frame(x = AAPL.dens$x, y = dnorm(AAPL.dens$x,mean(AAPL.ret$x),sd(AAPL.ret$x))), aes(x, y), stat = "identity", color = "red") +
  labs(title = "Returns Distribution",
       x = "Returns",
       y = "Density")
```

Using the methods of the workshop estimate parameters of normal model and robust normal model for the Apple returns.

Estimate realized arithmetic volatility of Apple returns using both normal and robust models.

Volatility is calculated as: $$\sigma=Sd\sqrt{365},$$ where $Sd$ is standard deviation of daily arithmetic returns $S(t)−S(t−1)$, $S(t)$ is stock price.

```{r}

```










