---
title: "Week 6: Homework Assignment"
author: "Joshua Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
output:
  html_document:
    theme: united
    df_print: paged
    highlight: textmate
    code_folding: show
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: inline
always_allow_html: yes
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp=.618, fig.align="center",
                      fig.path="Figs/",
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here, rstan, HDInterval, tictoc, furrr, latex2exp, quantmod)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())

# Parallel Stan
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

# Data 
Consider daily simple returns of AAPL, which, for example, can be obtained as:
```{r}
getSymbols("AAPL", from = "2015-1-1", to = "2015-12-31")
```

The data are in the file `AAPL_2015.csv` 
```{r}
AAPL.1.day <- as.matrix(AAPL)
AAPL.returns.2015 <- diff(AAPL.1.day[, 6])
```

```{r}
AAPL.ret <- read_csv("documents-MScA Bayesian Methods (32014)-MScA 32014 Lecture 7-AAPL_2015.csv")
(AAPL.ret)
```

Visualize the distribution of the returns.
```{r}
AAPL.dens <- density(AAPL.returns.2015)
data.frame(returns = AAPL.returns.2015) %>%
  ggplot() +
  geom_density(aes(returns)) +
  geom_line(data = data.frame(x = AAPL.dens$x, y = dnorm(AAPL.dens$x, mean(AAPL.ret$x), sd(AAPL.ret$x))), aes(x, y), stat = "identity", color = "red") +
  labs(title = "Returns Distribution",
       x = "Returns",
       y = "Density")
```

Using the methods of the workshop to estimate parameters of normal model and robust normal model for the Apple returns.
```{r}
returns <- AAPL.returns.2015
sd_returns <- sd(AAPL.returns.2015)
mean_returns <- mean(AAPL.returns.2015)
arithmetic_volatility <- sd_returns * sqrt(365)
Ntotal <- length(AAPL.returns.2015)

data = list(
  y = returns,
  Ntotal = Ntotal,
  mean_mu = mean_returns,
  sd_mu = sd_returns
)
```

Estimate realized arithmetic volatility of Apple returns using both normal and robust models.

Volatility is calculated as: $$\sigma=Sd\sqrt{365},$$ where $Sd$ is standard deviation of daily arithmetic returns $S(t)−S(t−1)$, $S(t)$ is stock price.

# Fit Normal Model
```{r Fit-model, message=FALSE, warning=FALSE}
fit <- stan(file = "model_1_week_6_assignment.stan", data = data, seed = 1)
```

##  Analyze convergence
```{r Show-fit}
plot(fit)
show(fit)
names(fit)
summary(fit)
```

Trace shows $\kappa$ not overlapping.
```{r Trace}
stan_trace(fit) +
  scale_color_viridis_d() + 
  theme(legend.position = "none")
```

```{r Auto-correlation}
stan_ac(fit, separate_chains = T)
```

```{r Parameters}
pairs(fit, pars = c("mu", "sigma"))
stan_hist(fit)
```

```{r Parameters}
stan_diag(fit, information = "treedepth")
```

# Fit Robust Normal (fat tails) Model
```{r Fit-robust-model, message=FALSE, warning=FALSE}
fit_robust <- stan(file = "robust_normal_model_week_6_assignment.stan", data = data, seed = 1)
```

##  Analyze convergence
```{r Show-fit-robust}
plot(fit_robust, pars = c("mu", "sigma"))
show(fit_robust)
names(fit_robust)
summary(fit_robust)
```

Trace shows $\kappa$ not overlapping.
```{r Trace}
stan_trace(fit_robust) +
  scale_color_viridis_d() + 
  theme(legend.position = "none")
```

Robust model has some auto-correlation.
```{r Auto-correlation}
stan_ac(fit_robust, separate_chains = T)
```

```{r Parameters}
pairs(fit_robust, pars = c("mu", "sigma"))
stan_hist(fit_robust)
```

```{r Parameters}
stan_diag(fit_robust, information = "treedepth")
```



