---
title: "Workshop 1"
author: "Josh Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
always_allow_html: yes
output:
  github_document: 
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/documents-MScA Bayesian Methods (32014)-MScA 32014 Lecture 1-3WayData.csv", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

# Joint Distribution: 3-Dimensional Case
```{r}
u <- c("u1", "u2")
v <- c("v1", "v2", "v3")
w <- c("w1", "w2", "w3")
matr.u0 <- paste("u0", outer(v, w, paste, sep = ","), sep = ",")
dim(matr.u0) <- c(3, 3)
matr.u1 <- paste("u1", outer(v, w, paste, sep = ","), sep = ",")
dim(matr.u1) <- c(3, 3)
matr.u0
```

```{r Read-data}
data3way <-
  read_delim("documents-MScA Bayesian Methods (32014)-MScA 32014 Lecture 1-3WayData.csv",
             delim = ",") %>%
  data.frame()
```

```{r}
head(data3way)
```

```{r}
mat.u0 <- table(subset(data3way, u == 0)[, 1], subset(data3way, u == 0)[, 2])
mat.u1 <- table(subset(data3way, u == 1)[, 1], subset(data3way, u == 1)[, 2])
```


```{r}
mat.u0
```

```{r}
idx.v1 <- data3way$v == 1
idx.w1 <- data3way$w == 1
idx.u1 <- data3way$u == 1
sum(idx.v1 * idx.w1 * idx.u1) # element (1, 1) of mat.u1
```

```{r}
idx.v2 <- data3way$v == 2
sum(idx.v2 * idx.w1 * idx.u1) # element (1, 2) of mat.u1
```

```{r}
colnames(mat.u1) <- colnames(mat.u0) <- c("v1", "v2", "v3")
rownames(mat.u1) <- rownames(mat.u0) <- c("w1", "w2", "w3")

data3way.array <-
  array(rep(NA, 18),
        dim = c(3, 3, 2),
        dimnames = list(
          paste("w", 1:3, sep = ""),
          paste("v", 1:3, sep =
                  ""),
          paste("u", 0:1, sep =
                  "")
        ))

data3way.array[, , 1] <- mat.u0
data3way.array[, , 2] <- mat.u1
data3way.array
```

```{r}
N <- sum(data3way.array)
(data3way.array.p <- data3way.array / N)
```

## 1.2 Marginal Distributions
Create marginal distribution for `u` as vector `uMarginal`.
```{r}
(uMarginal <- apply(data3way.array.p, 3, sum))
```

Create marginal distribution for `v` as vector `vMarginal`.
```{r}
(vMarginal <- apply(data3way.array.p, 2, sum))
vMarginal["v1"]
```

Create marginal distribution for `w` as vector `wMarginal`.
```{r}
(wMarginal <- apply(data3way.array.p, 1, sum))
wMarginal["w1"]
```

## 1.3 Conditional Distributions
Create conditional distribution $p(w,v|u=1)$ as matrix `cond.v.w.given.u1`.
```{r}
cond.v.w.given.u1 <- data3way.array.p[, , "u1"] / sum(data3way.array.p[, , "u1"])
cond.v.w.given.u1["w1", ]
sum(cond.v.w.given.u1) # check
```

Create conditional distribution $p(v|u=1)$ as vector `cond.v.given.u1`.
```{r}
cond.v.given.u1 <- apply(data3way.array.p[, , "u1"], 2, sum) / sum(data3way.array.p[, , "u1"])
cond.v.given.u1["v1"]
sum(cond.v.given.u1) # check
```

Create conditional distribution $p(w|v=2,u=1)$ as vector `cond.w.given.u1.v2`.
```{r}
cond.w.given.u1.v2 <- data3way.array.p[, "v2", "u1"] / sum(data3way.array.p[, "v2", "u1"])
cond.w.given.u1.v2["w1"]
sum(cond.w.given.u1.v2) # check
```

Compare the vectors $p(w|v2,u1)p(v2|u1)p(u1)$ and $p(w,v,u)[,v2,u1]$
```{r}
rbind(uMarginal["u1"]*cond.v.given.u1["v2"]*cond.w.given.u1.v2,data3way.array.p[,"v2","u1"])
```

# 2 Simulation Using Conditional Distributions
Let the marginal distribution for random variable `u` be Bernoulli with $p(u=0)=0.55$, $p(u=1)=0.45$.
Let conditional distributions for random variables $(v|u=0)$ and $(v|u=1)$, taking values 1, 2, 3 be
```{r}
(pCond.v.given.u0 <- c(.7, .2, .1))
(pCond.v.given.u1 <- c(.1,.2,.7))
```

Let random variable $(w|v,u)$ take values $1,2,3$ with probabilities $p(w|v,u)$, given by the following:
```{r}
p.given.u0.v1 <- c(.3, .3, .4)
p.given.u0.v2 <- c(.5, .3, .2)
p.given.u0.v3 <- c(.6, .2, .2)
p.given.u1.v1 <- c(.2, .3, .5)
p.given.u1.v2 <- c(.2, .2, .6)
p.given.u1.v3 <- c(.1, .7, .2)
```

Simulate joint sample $(w,v,u)$ of lenth $n=500$.
Use `set.seed(11)` Start with simulation of u.
For each simulated value `u` generate `v` from the corresponding conditional distribution $p(v|u)$.
Finally, for each pair $v,u$ simulate `w` from $p(w|v,u)$.
```{r}
v_probs <- function(u) {
  switch(u + 1, 
         pCond.v.given.u0,
         pCond.v.given.u1)
}

w_probs <- function(u, v) {
  switch(u + 1,
         switch(v,
                p.given.u0.v1,
                p.given.u0.v2,
                p.given.u0.v3),
         switch(v,
                p.given.u1.v1,
                p.given.u1.v2,
                p.given.u1.v3)
         )
}

set.seed(11)
sim.u <- rbinom(500, 1, prob = .45)
sim.v <- map_dbl(map(sim.u, ~ v_probs(.x)), ~ sample(1:3, 1, replace = TRUE, prob = .x))
sim.w <- map_dbl(map2(sim.u, sim.v, ~ w_probs(.x, .y)), ~ sample(1:3, 1, replace = TRUE, prob = .x))

head(cbind(sim.w, sim.v, sim.u), 25)
```

# 3 Conditional Expected Values
Calculate unconditional expected value $E[v]$. Random variable `v` can take values $1,2,3$ with corresponding probabilities.
```{r}
vMarginal
```

Then the unconditional mean value is
```{r}
c(1, 2, 3) %*% vMarginal
```

Calculate conditional expected value $E[v|u]$.

First, find conditional mean values $E[v|u=u0]=E[v|u0]$ and $E[v|u=u1]=E[v|u1]$.

The random variable $(v|u0)$ takes values $1,2,3$ with corresponding probabilities $p(v=1|u0)$, $p(v=2|u0)$, $p(v=3|u0)$, given by the vector.
```{r}
(cond.v.given.u0 <- apply(data3way.array.p[, , "u0"], 2, sum) / uMarginal["u0"])
```

Taking conditional expected value with respect to this distribution $E[v|u0]$ is:
```{r}
(exp.v.given.u0 <- c(1, 2, 3) %*% cond.v.given.u0)
```

The random variable $(v|u1)$ takes the same values $1,2,3$, but with different probabilities $p(v|u1)$:
```{r}
cond.v.given.u1
```

Thus conditional expected value $E[v|u1]$ is
```{r}
(exp.v.given.u1 <- c(1, 2, 3) %*% cond.v.given.u1)
```

Note that the conditional expected value $E[v|u]$ takes two different values: $E[v|u0]$ and $E[v|u1]$ with probabilities $p(v|u0)$ and $p(v|u1)$, correspondingly.

This means that $E[v|u]$ is a random variable and its (unconditional) expected value is $$E[v|u]=E[v|u0]p(u=u0)+E[v|u1]p(u=u1)$$ 

Calculate this unconditional expected value:
```{r}
(uncond.exp.v.given.u <-
   exp.v.given.u0 * uMarginal["u0"] + exp.v.given.u1 * uMarginal["u1"])
```





