---
title: "Workshop 6.2"
author: "Joshua Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
always_allow_html: yes
output:
  github_document: 
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

# 2 Review of MCMC: binomial model

## 2.1 Data
Create vector of zeros and ones of length $k=20$ and probability of success $\theta=0.6$
```{r}
set.seed(9374)
Y <- rbinom(20, size = 1, p = .6)
```

## 2.2 Running MCMC
Recall that Metropolis-Hastings MCMC algorithm consists of the following steps:

### 2.2.1 Generate new proposal using some convenient distribution.
Typically new proposals are generated from Gaussian distribution centered at the current value of Markov chain with sufficiently small variance. Write a function for new proposals `proposalFun <- function(oldTheta)` which simulates newTheta from Gaussian distribution with mean value oldTheta and fixed standard deviation of small enough level to generate reasonable proposals for $\theta$ between 0 and 1.

It is recommended to truncate the proposed values between extreme levels of, for example, 0.01 and 0.99 to prevent values exactly at 0 and 1.
```{r}
generate_proposal <- function(oldTheta, tuneSigma = .1) {
  newTheta <- rnorm(1, mean = oldTheta, sd = tuneSigma)
  newTheta <- min(.99, newTheta)
  newTheta <- max(.01, newTheta)
  newTheta
}
```

### 2.2.2 Decide whether the proposal `newTheta` is accepted or rejected.
```{r}
binomialLikelihood <- function(Theta, mySample) {
  prod(dbinom(mySample, 1, Theta))
}

binomialPrior <- function(Theta, A, B) {
  dbeta(Theta, A, B)
}
```

```{r}
metropolisRule <- function(thetaOld, thetaNew, mySample, funLikelihood, funPrior) {
  dec_value <- (funLikelihood(thetaNew, mySample) * funPrior(thetaNew, priorParam[[1]], priorParam[[2]])) /
    (funLikelihood(thetaOld, mySample) * funPrior(thetaOld, priorParam[[1]], priorParam[[2]]))
  
  dec_value <- min(1, dec_value)
  
  if(runif(1) <= dec_value) {
    acceptance <- FALSE
  } else {
    acceptance <- TRUE
  }
  acceptance
}
```

### 2.2.3 Initiate parameters
When all functions are prepared initiate necessary parameters:
```{r}
# Initiate
chainMax <- 500               # length of Markov chain
Chain <- numeric(0)           # empty vector for the chain
Rejected <- numeric(0)        # empty vector for rejected values
chainLength <- length(Chain)  # current length of chain
tuneSigma <- .1               # standard deviation of proposal distribution
priorParam <- c(A = 10, B = 10)    # parameters of prior beta distribution
ThetaOld <- .5                # initial Theta
```

### 2.2.4 Run MCMC
Run loop until number of accepted values in Chain reaches `chainMax`.
In this loop:

* Generate new proposal `ThetaNew`

* Apply Metropolis decision rule. If new proposal is accepted add it to `Chain` and increase `chainLength`, then reset `ThetaOld` to `ThetaNew`. If new proposal is rejected then add it to vector `Rejected`

```{r}
while(chainLength < 500) {
  Chain <- numeric(0)           # empty vector for the chain
  Rejected <- numeric(0)        # empty vector for rejected values
  i <- 1
  
  new_proposal <- generate_proposal(ThetaOld, tuneSigma = tuneSigma)
  accept <- metropolisRule(thetaOld = ThetaOld, new_proposal, Y, binomialLikelihood, binomialPrior)
  
  if (accept) {
    Chain[i] <- new_proposal
  } else {
    Rejected[i] <- new_proposal
  }
  
  i <- i + 1
  chainLength <- length(Chain)
  
}
```







