---
title: "Workshop 2"
author: "Joshua Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
always_allow_html: yes
output:
  github_document: 
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())
source("documents-MScA Bayesian Methods (32014)-KruschkeCode-DBDA2E-utilities.R")
source("documents-MScA Bayesian Methods (32014)-KruschkeCode-BernBeta.R")
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/documents-MScA Bayesian Methods (32014)-KruschkeCode-BernBeta.R", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

# 3 Example: Mammography Screening Controvercy
## 3.1 Problem
According to CDC, probability of breast cancer among women of age 50-60 years old is 2.28%.
Screening using mammography has sensitivity (i.e. probability of detecting the disease correctly) 75%-94%. Specificity of the test (i.e. probability of correct "no disease" conclusion) is 83%-98%. Assume the most optimistic parameters of the test (94% and 98%, correspondingly for sensitivity and specificity) What is the probability that randomly selected woman with positive test result has the disease?
```{r}
# Prior p(theta1)
p.theta1 <- .0228
# Sensitivity: p(tau1|theta1)
p.tau1.theta1 <- .94
# Specificity: p(tau0|theta0)
p.tau0.theta0 <- .98
# Complimentary probabilities
p.theta0 <- 1 - p.theta1
p.tau1.theta0 <- 1 - p.tau0.theta0
# Bayes
(p.theta1.tau1 <- p.tau1.theta1 * p.theta1 / (p.tau1.theta1 * p.theta1 + p.tau1.theta0 * p.theta0))
```

Assume that randomly selected woman tested positive and then retested with negative test result.
After both tests what is the probability of disease?
```{r}
# Set the prior to the new probability of having the disease:
pDisease <- p.theta1.tau1
# Bayes rule for second, negative test:
(p.theta1.tau0 <- ((1.0 - p.tau1.theta1) * pDisease / + ((1.0 - p.tau1.theta1) * pDisease + (1.0 - p.tau1.theta0) * (1.0 - pDisease))))
```

# 4 Example of Using Bayes Theorem in Statistics
```{r}
(Theta <- seq(0 , 1 , length = 5))  # Sparse teeth for Theta.

pTheta <- pmin( Theta , 1-Theta ) # Triangular shape for pTheta.
(pTheta <- pTheta/sum(pTheta)) # Make pTheta sum to 1.0
Data  <- c(rep(0, 0), rep(1, 1))      # Single flip with 1 head
```

## 4.1 Binomial model, triangular prior with 3 values

## 4.2 Binomial model, triangular prior with 11 values

## 4.3 Binomial model, uniform prior with 1001 values
```{r}
Theta <- seq( 0 , 1 , length = 1001 ) # fine grid for Theta.
pTheta <- rep(1,length(Theta))      # Uniform (horizontal) shape for pTheta.
pTheta <- pTheta/sum(pTheta)        # Make pTheta sum to 1.0
Data <- c(rep(0,0),rep(1,1))        # Single flip with 1 head
```

```{r}

```

















