---
title: "Course Project"
author: "Joshua Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
always_allow_html: yes
output:
  html_document:
    theme: united
    df_print: paged
    highlight: textmate
    code_folding: show
    toc: true
    toc_float: true
  github_document:
editor_options: 
  chunk_output_type: inline
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp=0.618, fig.path='Figs/',
                      warning=FALSE, message=FALSE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here, shinystan, rstan, HDInterval) 

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

# Data
The data of the project contains 5 factors:

* `Sex` (2 levels)
* `Race` (4 levels)
* `Age` (5 levels)
* `Education` (3 levels)
* `State` (51 levels)

The total number of combinations of levels is 6120.

Look at the data.

Check the variables notation:

```{r}
dat <- read.csv("assignments-Bayesian Methods-Course Project 32014-MScA_32014_BayesianMethods_CourseProjectData.csv")
head(dat)
```

```{r}
unique(dat$sex)
unique(dat$race)
unique(dat$age)
unique(dat$education)
unique(dat$state)
```

# Model
After running MCMC with the these data and the model obtain a Markov chain posterior sample for 870 parameters including 2-way interactions.
Each Markov chain of the stan object `obama_fit` has length 36000.

Explore the fitted model object.
It is not necessary to reproduce the results shown below, but if you see significant differences, please, report.

Launch `shinystan()` to explore the chains for convergence.

```{r}
load("assignments-Bayesian Methods-Course Project 32014-Run05272016-fit_ext_20160527_103144.Rdata")
```

```{r eval=FALSE}
launch_shinystan(obama_fit)
```

# Analysis
## Extract chains for further analysis
```{r}
MCMC <- rstan::extract(obama_fit)
MCMC %>% names()
```

## Significant parameters
Explore `sum.obama_fit` - summary of `obama_fit`.

Find parameters which are significantly different from zero: zero does not belong to 95% HDI.
Show selected parameters.
```{r}
sum.obama_fit <- summary(obama_fit)$summary %>% data.frame()

significant_co <- function(stan_fit) {
  summ <- summary(stan_fit)$summary %>% data.frame()
  !(summ$X2.5. < 0 & summ$X97.5. > 0)
}
```

```{r}
sum.obama_fit[significant_co(obama_fit), c(4, 8)]
```

The model parameters are re-centered to satisfy additional constraint $(\sum_i \beta_i = 0)$, so, for example slopes of gender predictor are:

```{r}
sum.obama_fit[2:3, 1]
sum(sum.obama_fit[2:3, 1])
```

As a result estimated intercept is close to 0.5:

```{r}
plot(obama_fit, pars = c("b_0"))
hist(MCMC$b_0)
mean(MCMC$b_0)
```

## Reviewing main effects
Plot HDIs for the 5 main parameters.
For example:
```{r}
plot(obama_fit, pars = c("b_sex"))
```

## Interactions
Interpretation of interactions.
Look, for example, at interaction between sex and race.

```{r}
plot(obama_fit, pars = c("b_sex_race"))
```

```{r}
sum.obama_fit[67:74, c(1, 4, 8)]
```

Two of the coefficients are significantly different from zero:
```{r}
sum.obama_fit[c(67, 70), c(1, 4, 8)]
```

Recall that `sex[1]` is "Male" and `sex[2]` is "Female", and `race[1]` means "White" and `race[4]` means "Other".
Main parameters and their corresponding interactions are then:

```{r}
sum.obama_fit[c(1:4, 7, 67, 70), c(1, 4, 8)]
```

Interpret these parameters as influence over odds ratio.
```{r}
(odds <- exp(sum.obama_fit[c(1:4, 7, 67, 70), c(1, 4, 8)][, 1]))
```

The baseline odds ratio is 1.643976, meaning that the ratio of people approving Obama as candidate to disapproving him is 1.643976.

```{r}
prod(odds[1:2])
```

And among females it is higher:
```{r}
prod(odds[c(1, 3)])
```

Approval odds for 1.Whiteand 4.Other are:
```{r}
odds[c(1, 4)]
prod(odds[c(1, 4)])
odds[c(1, 5)]
prod(odds[c(1, 5)])
```

Approval odds among white males $(sex=1,race=1)$ is:
```{r}
odds[c(1, 2, 4, 6)]
prod(odds[c(1, 2, 4, 6)])
```

Which is better than in the case when interaction is not included:
```{r}
odds[c(1, 2, 4)]
prod(odds[c(1, 2, 4)])
```

Note that since interaction `b_sex_race[2,1]` is not significantly different from zero, odds of approval among white women may not be significantly different from
```{r}
odds[c(1, 3, 4)]
prod(odds[c(1, 3, 4)])
```

# Project questions

1. Find the most supportive groups corresponding to each of the main effects for Barack Obama in 2012. 

1. Find the least supportive group corresponding to each of the main effects.

1. Compare odds of approval by males grouped by race and education. Which of such subgroups shows the highest and the lowest support for the candidate.

1. Answer the same question, but for females

For convenience create groups of necessary parameters and give them intuitive column names.

```{r}
ch_b0 <- MCMC$b_0
ch_b_sex <- MCMC$b_sex
colnames(ch_b_sex) <- levels(dat$sex)

ch_b_race <- MCMC$b_race
colnames(ch_b_race) <- levels(dat$race)

ch_b_age <- MCMC$b_age
colnames(ch_b_age) <- levels(dat$age)

ch_b_education <- MCMC$b_education
colnames(ch_b_education) <- levels(dat$education)

ch_b_state <- MCMC$b_state
colnames(ch_b_state) <- levels(dat$state)

ch_b_sex_race <- MCMC$b_sex_race
dimnames(ch_b_sex_race)[[2]] <- levels(dat$sex)
dimnames(ch_b_sex_race)[[3]] <- levels(dat$race)

ch_b_sex_education <- MCMC$b_sex_education
dimnames(ch_b_sex_education)[[2]] <- levels(dat$sex)
dimnames(ch_b_sex_education)[[3]] <- levels(dat$education)

ch_b_race_education <- MCMC$b_race_education
dimnames(ch_b_race_education)[[2]] <- levels(dat$race)
dimnames(ch_b_race_education)[[3]] <- levels(dat$education)
```

## Main effects
### Grouped by gender
Calculate mean, lower and upper HDI levels for odds of approval by gender.
Calculate mean values and HDI limits for log-odds and then take exponents of them.

```{r}
calculate_odds <- function(var) {
  rbind(mean = apply(var, 2, function(x) exp(mean(x + ch_b0))), apply(var, 2, function(x) exp(hdi(x + ch_b0))))
}
```

### Grouped by sex
```{r}
(ch_b_sex_odds <- calculate_odds(ch_b_sex))
which.max(ch_b_sex_odds[1, ])
which.min(ch_b_sex_odds[1, ])
```

Define convenience function for plotting:
```{r}
add_spaces <- function(string) {
  str_remove_all(string, "\\d\\.") %>% 
      str_replace_all("([[:upper:]])", " \\1") %>% 
      str_trim("both")
} 

plot_odds <- function(odds_matrix) {
  data <- odds_matrix[1,] %>%
    data.frame(odds = .) %>%
    rownames_to_column(var = "var")
  
  edu_check <- unique(data$var %>% str_detect("College"))
  
  if (edu_check) {
    edu_names <- add_spaces(data$var)
  }
  
  data <- data %>% 
    mutate(var = str_remove_all(var, "\\d\\.")) %>% 
    mutate(var = if(edu_check) edu_names else var,
           var = factor(var, levels = var[order(odds, decreasing = TRUE)], ordered = TRUE))
  age_check <- unique(data$var %>% str_detect("\\d"))
  data %>% 
    ggplot(aes(var, odds), odds) +
    geom_col(fill = "skyblue") +
    labs(
      title = glue::glue("{levels(data$var)[1]}{ifelse(age_check, ' Age Group', '')} Shows Highest Support for Obama"),
      x = "Group",
      y = "Odds"
    )
}
```

```{r}
plot_odds(ch_b_sex_odds)
```

### Grouped by race
```{r}
(ch_b_race_odds <- calculate_odds(ch_b_race))
which.max(ch_b_race_odds[1, ])
which.min(ch_b_race_odds[1, ])
```

```{r}
plot_odds(ch_b_race_odds)
```

### Grouped by age
```{r}
(ch_b_age_odds <- calculate_odds(ch_b_age))
which.max(ch_b_age_odds[1, ])
which.min(ch_b_age_odds[1, ])
```

```{r}
plot_odds(ch_b_age_odds)
```

### Grouped by education
```{r}
(ch_b_education_odds <- calculate_odds(ch_b_education))
which.max(ch_b_education_odds[1, ])
which.min(ch_b_education_odds[1, ])
```

```{r}
plot_odds(ch_b_education_odds)
```

### Grouped by state
```{r}
(ch_b_state_odds <- calculate_odds(ch_b_state))
which.max(ch_b_state_odds[1, ])
which.min(ch_b_state_odds[1, ])
```

```{r}
plot_odds(ch_b_state_odds) +
  coord_flip()
```

## Interactions
### Males

Calculate mean values and log-odds limits for each race-education combination. Then take exponents of them.

Calculate mean and HDI of each column of the cube.

Compare groups.

```{r}
run <- 12
ch_bo_cube <- purrr::rerun(run, cbind(ch_b0)) %>% 
  bind_cols() %>% 
  set_names(rep("ch_b0", run)) %>% 
  as.matrix()
dim(ch_bo_cube) <- dim(ch_b_race_education)

ch_race_cube <- cbind(ch_b_race, ch_b_race, ch_b_race)
dim(ch_race_cube) <- dim(ch_b_race_education)
dimnames(ch_race_cube) <- dimnames(ch_b_race_education)

ch_b_education_cube <- cbind(ch_b_education, ch_b_education, ch_b_education, ch_b_education)

dim(ch_b_education_cube) <- dim(ch_b_race_education)[c(1, 3, 2)]

ch_b_education_cube <- aperm(ch_b_education_cube, c(1, 3, 2))

dimnames(ch_b_education_cube) <- dimnames(ch_b_race_education)

total_male <- ch_bo_cube + ch_b_sex[, "1.Male"] + ch_race_cube + ch_b_education_cube + ch_b_race_education
total_female <- ch_bo_cube + ch_b_sex[, "2.Female"] + ch_race_cube + ch_b_education_cube + ch_b_race_education
```

```{r}
(orM_mean <- apply(total_male, c(2, 3), function(x) mean(x)))
(orM_HDI <- apply(total_male, c(2, 3), function(x) hdi(x)))
exp(orM_mean)
exp(orM_HDI)
```

Define convenience functions for plotting:
```{r}
data_frame_interaction <- function(interaction_matrix) {
  interaction_matrix %>% 
  data.frame() %>% 
  rownames_to_column(var = "race") %>% 
  gather(key = "education", value = "odds", -1) %>% 
  mutate(race = str_remove_all(race, "\\d\\."),
         education = str_remove_all(education, "^X\\d\\.") %>% 
           add_spaces())
}

plot_interaction_odds <- function(interaction_matrix, title = NULL, subtitle = NULL) {
  data_frame_interaction(interaction_matrix) %>%
    group_by(race) %>%
    arrange(desc(odds), .by_group = TRUE) %>%
    add_count(race) %>%
    mutate(ordered_groups = factor(rep(1:n[[1]]), ordered = TRUE)) %>%
    ungroup() %>%
    ggplot(aes(race, odds, fill = education, group = ordered_groups)) +
    geom_col(position = "dodge") +
    geom_label(
      aes(label = round(odds, 2), group = ordered_groups),
      fill = "white",
      position = position_dodge(width = .9),
      size = 2
    ) +
    scale_fill_viridis_d(name = NULL) +
    labs(title = title,
         subtitle = subtitle,
         caption = "Note: Values are odds of supporting Obama.",
         x = "Race",
         y = "Odds") +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "top"
    )
}

```

```{r fig.asp=.80}
plot_interaction_odds(orM_mean, title = "Black Males with Some College Show Highest Support for Obama",
                      subtitle = "White Males with No College Show Least Support")
```

### Females
Calculate mean and HDI of each column of the cube.

Compare groups.
```{r}
(orF_mean <- apply(total_female, c(2, 3), function(x) mean(x)))
(orF_HDI <- apply(total_female, c(2, 3), function(x) hdi(x)))
exp(orF_mean)
exp(orF_HDI)
```

```{r fig.asp=.80}
plot_interaction_odds(orF_mean, 
                      title = "Black Females with Some College Also Show Highest Support for Obama",
                      subtitle = "White Females with No College Show Least Support")
```






