---
title: "Week 1 Assignment: Study on Smokers"
author: "Josh Goldberg"
date: "`r format(Sys.time(), '%B, %d %Y')`"
output:
  html_document:
    df_print: paged
  github_document: null
  pdf_document: default
editor_options:
  chunk_output_type: inline
always_allow_html: yes
---

```{r Global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
                      warning=FALSE, message=FALSE, cache=TRUE)
```

```{r Preamble, echo=FALSE}
# Enter package in p_load()
# If package is not installed, p_load() will install and load the package
if(!"pacman" %in% rownames(installed.packages())) {
  install.packages("pacman")
  }
pacman::p_load(tidyverse, ggthemes, here, faraway, janitor)

# Set default ggplot theme to tufte
theme_set(ggthemes::theme_tufte())
```

```{r Copy-files, echo=FALSE, eval=FALSE}
# Enter files to load to project directory in from = "~/Downloads/your_file_name_here"
file.copy(from = "~/Downloads/", to = here::here(), 
          overwrite = TRUE, recursive = FALSE, 
          copy.mode = TRUE)
```

**This project helps understanding conditional distributions**

The following data set is in `library(faraway)`.
In it participating women are categorized into groups of smokers or not, 7 age groups and groups of dead or alive after 20 years of study.
```{r Data}
data(femsmoke)
femsmoke
```

# 1 Joint distribution
Create joint distribution of 3 variables:

    * v - "smoker yes"=1 or "smoker no"=0
    * u - "dead yes"=1 or "dead no"=0
    * w - age category

$w=1$: 18-24

$w=2$: 25-34

$w=3$: 35-44

$w=4$: 45-54

$w=5$: 55-64

$w=6$: 65-74

$w=7$: 75+

```{r Functions}
make_array <- function() {
  array(rep(NA, 28),
        dim = c(length(unique(femsmoke$smoker)), length(unique(femsmoke$age)), 2),
        dimnames = list(
          paste("smoker", c("yes", "no"), sep = "."),
          paste("age", 1:7, sep = "."),
          paste("dead", c("yes", "no"), sep = ".")
        ))
}

make_counts <- function(x, count_col = y, df = femsmoke) {
  x <- enquo(x)
  count_col <- enquo(count_col)
  map2(
      df %>% pull(!! count_col),
      df %>% pull(!! x),
      ~ replicate(n = .x, .y)
    ) %>% 
    unlist()
}
```

```{r Joint-probs-1}
dead_yes <- matrix(NA, nrow = length(unique(femsmoke$smoker)), ncol = length(unique(femsmoke$age)))
dead_no <- matrix(NA, nrow = length(unique(femsmoke$smoker)), ncol = length(unique(femsmoke$age)))

colnames(dead_yes) <- colnames(dead_no) <- unique(femsmoke$age)
rownames(dead_yes) <- rownames(dead_no) <- unique(femsmoke$smoker)

w <- unique(femsmoke$age)
v <- unique(femsmoke$smoker)
u <- unique(femsmoke$dead)

pwalk(list(
  row = match(femsmoke$smoker, v),
  col = match(femsmoke$age, w),
  y = femsmoke$y,
  dead = femsmoke$dead
), function(row, col, y, dead) {
  switch(dead,
         yes = dead_yes[row, col] <<- y,
         no = dead_no[row, col] <<- y)
})

femsmoke_array <- make_array()

femsmoke_array[, , "dead.yes"] <- dead_yes
femsmoke_array[, , "dead.no"] <- dead_no

total_count <- sum(femsmoke_array)

(femsmoke_joint_p <- femsmoke_array / total_count)
sum(femsmoke_joint_p)
```

```{r Joint-probs-2}
new_femsmoke <- tibble(age = make_counts(age),
                       dead = make_counts(dead),
                       smoker = make_counts(smoker))

femsmoke_mod_u1 <- t(table(new_femsmoke %>% filter(dead == "yes") %>% pull(age), 
                         new_femsmoke %>% filter(dead == "yes") %>% pull(smoker)))

femsmoke_mod_u0 <- t(table(new_femsmoke %>% filter(dead == "no") %>% pull(age), 
                         new_femsmoke %>% filter(dead == "no") %>% pull(smoker)))

femsmoke_array <- make_array()

femsmoke_array[, , "dead.yes"] <- femsmoke_mod_u1
femsmoke_array[, , "dead.no"] <- femsmoke_mod_u0

total_count <- sum(femsmoke_array)

(femsmoke_joint_p <- femsmoke_array / total_count)
sum(femsmoke_joint_p)
```

```{r Joint-probs-3}
femsmoke_joint_p_xtabs <- xtabs(y ~ smoker + age + dead, data = femsmoke) / sum(femsmoke$y)
sum(femsmoke_joint_p_xtabs)
```

# 2 Marginal distributions
Create marginal distribution for *u* (dead).
```{r Marginal-u}
(uMarginal_dead <- apply(femsmoke_joint_p, 3, sum))
sum(uMarginal_dead)
```

Create marginal distribution for *v* (smoke).
```{r Marginal-v}
(vMarginal_smoke <- apply(femsmoke_joint_p, 1, sum))
sum(vMarginal_smoke)
```

Create marginal distribution for *w* (age).
```{r Marginal-w}
(wMarginal_age <- apply(femsmoke_joint_p, 2, sum))
sum(wMarginal_age)
```

# 3 Conditional distributions
Create conditional distribution $p(w,v|u="alive")=p(smoke,age|alive)$.
```{r Conditional-smoke-age-cond-alive}
(cond_v_w_given_uAlive <- femsmoke_joint_p[, , "dead.no"] / uMarginal_dead["dead.no"])
sum(cond_v_w_given_uAlive)
```

Create conditional distribution $p(v|u="alive")=p(smoke|alive)$.
```{r Conditional-smoke-cond-alive}
(cond_v_given_uAlive <- apply(femsmoke_joint_p[, , "dead.no"], 1, sum) / uMarginal_dead["dead.no"])
sum(cond_v_given_uAlive)
```

Create conditional distribution $p(w|u="alive",v="smoker")=p(age|alive,smoke)$.
```{r Conditional-age-cond-alive-smoke}
(cond_w_given_uAlive_vSmoke <- femsmoke_joint_p["smoker.yes", , "dead.no"] / sum(femsmoke_joint_p["smoker.yes", , "dead.no"]))

sum(cond_w_given_uAlive_vSmoke)
```

Compare the vectors $p(w|v2,u1)p(v2|u1)p(u1) and p(w,v,u)[,v2,u1]$.
```{r Compare-vectors}
rbind(
  uMarginal_dead["dead.no"] * cond_v_given_uAlive["smoker.yes"] * cond_w_given_uAlive_vSmoke,
  t(femsmoke_joint_p[, 1:7, "dead.no"])[, 1]
)
```

# 4 Simulation
Let the marginal distribution for age group be $p(age)=p(w)$ estimated marginal distribution from the sample:
```{r Marginal-sample}
wMarginal_age
```

Given simulated age group, simulate variable `v = smoke` using conditional distribution $p(v|w)$, i.e. using probabilities $p(smoke.yes|age)$, $p(smoke.no|age)$.

Given simulated variables `w` for age and `v` for smoke, simulate mortality variable `u` using distribution $p(dead|v,w)$, $p(alive|v,w)$.

Using the described procedure simulate outcomes for 100 participants.
Use seed `set.seed(284)` for reproducibility.
```{r Simulate-1}
set.seed(284)
sim_age <- sample(1:7, 100, replace = TRUE, prob = wMarginal_age)

smokers <- apply(femsmoke_joint_p["smoker.yes", , ], 1, sum) / wMarginal_age
non_smokers <- apply(femsmoke_joint_p["smoker.no", , ], 1, sum) / wMarginal_age

conditional_smoke_age <- rbind(smoker.yes = smokers, smoker.no = non_smokers)

sim_smoke <- map_dbl(sim_age, ~ sample(0:1, 1, replace = TRUE, prob = c(non_smokers[.x], smokers[.x])))

smokers_mort <- femsmoke_joint_p["smoker.yes", , ] / (conditional_smoke_age["smoker.yes", ] * wMarginal_age)
non_smokers_mort <- femsmoke_joint_p["smoker.no", , ] / (conditional_smoke_age["smoker.no", ] * wMarginal_age)

sim_dead <- pmap_dbl(list(sim_age, sim_smoke), ~ sample(1:0, 1, replace = TRUE, 
                                          prob = switch(.y + 1,
                                                        non_smokers_mort[.x, ],
                                                        smokers_mort[.x, ])))

attempt <- bind_cols(ages = sim_age, smokers = sim_smoke, dead = sim_dead) %>% 
  head(25)

answer <- tribble(
  ~ages, ~smokers, ~dead,
    3,       0,      0,
    6,       0,      0,
    3,       0,      0,
    3,       0,      0,
    4,       1,      0,
    2,       1,      0,
    7,       0,      1,
    5,       0,      0,
    5,       1,      1,
    5,       1,      0,
    5,       1,      1,
    3,       1,      0,
    2,       0,      0,
    7,       1,      1,
    3,       0,      0,
    1,       0,      0,
    6,       0,      1,
    2,       0,      0,
    5,       1,      0,
    2,       0,      0,
    4,       0,      0,
    7,       1,      1,
    3,       1,      0,
    3,       0,      0,
    1,       0,      0,
)

map2(attempt, answer, ~ which(.x != .y))
```

